---
title: 'PSTAT 174 HW #4'
author: "Chris Meade"
date: "2/14/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=T)
```

## Problem 2

Let $\{X_t\}$ be a $MA(1)$ process such that $X_t=Z_t+\theta Z_{t-2}$. Assume that $\{X_t\}$ has mean $\mu$, $\theta=-0.6$, $\sigma^2=1$, and $\bar{x}_{100}=0.157$. Since $\{X_t\}$ is a $MA(1)$ process, we have 

$$\gamma_{X}(h)=\sigma^2(1+\theta^2)\ \text{if}\ h=0$$
$$\gamma_{X}(h)=\sigma^2\theta\ \text{if}\ h=1,-1$$
$$\gamma_{X}(h)=0\ \text{if}\ |h|>1$$

Since $n=100$ is sufficiently large, we can estimate the variance of $\{X_t\}$ with $n^{-1}v$ where $$
\begin{aligned}
v &= \gamma_X(0)+2\sum_{1\leq h \leq n} ((1-\dfrac{h}{n})\gamma_X(h))\\
&=\sigma^2(1+\theta^2)+2((1-\dfrac{1}{100})\theta)\\
&=\sigma^2+\sigma^2\theta^2+2\theta-2\theta\dfrac{1}{100}
\end{aligned}$$

Then $$n^{-1}v=100^{-1}(\sigma^2+\sigma^2\theta^2+2\theta-2\theta\dfrac{1}{100})$$

Plugging in estimate for $\sigma^2$ and $\theta$ yields

$$n^{-1}v=\dfrac{0.172}{100}=0.00172$$

Since $X_t$ is approximately normal distributed, we may we the $Z$ test statistic in the construction of our confidence interval.

We now construct the confidence interval $$\begin{aligned}
&I=(\bar x_{100}-Z_{.025}*0.00172, \bar x_{100}+Z_{.025}*0.00172)\\
&=(0.157-1.96*0.00172, 0.157+1.90*0.00172)\\
&=(0.153629,0.160268)
\end{aligned}$$

Since $0\not\in I$, we would reject the hypothesis that $\mu=0$.

## Problem 3

Using Bartlet's formula, we cacluate the diagonal entries of the matrix $W$, $w_{1,1}$ and $w_{2,2}$.

$$w_{1,1}=1-3\rho(1)^2+4\rho(1)^4$$
and
$$w_{2,2}=1+2\rho(1)^2$$

To contruct the $95\%$ confidence interval for $\rho(1)$, we plug our estimates of $\rho(1)$ and $\rho(2)$ into the equation for $w_{1,1}$.

$$\begin{aligned}
\hat w_{1,1}&=1-3\rho(1)^2+4\rho(1)^4 \\
&=1-3(0.438)^2+4(0.145)^4 \\
&=0.426
\end{aligned}
$$

The confidence interval of $\rho(1)$ is of the form $(\hat\rho(1)-Z_{0.025}\sqrt{w_{1,1}/n},\hat\rho(1)+Z_{0.025}\sqrt{w_{1,1}/n})$.
Substitution our estimates yields $(0.310,0.566)$.

We now repeat the process to construct a $95\%$ confidence interval for $\rho(2)$.

$$\begin{aligned}
\hat w_{2,2}&=1+2\rho(1)^2 \\
&=1+2(0.438)^2 \\
&=1.387
\end{aligned}
$$

Then the confidence interval for $$\rho(2)=(\hat\rho(2)-Z_{0.025}\sqrt(w_{2,2}/100),\hat\rho(2)+Z_{0.025}\sqrt(w_{2,2}/100))=(-0.085,0.376)$$

If $\theta=0.6$, then $\rho(1)=\dfrac{0.6}{1+0.6^2}=0.441$. Since this value in an element of the confidence interval for $\rho(1)$, and because the confidence interval for $\rho(2)$ contains $0$, we fail to reject the hypothesis that $\theta=0.6$.

## Problem 4

This dataset contains the average monthly atomospheric $C0_2$ concentration. The data is available from https://www.esrl.noaa.gov/gmd/ccgg/trends/data.html. This data is important due to the relationship between atmospheric $CO_2$ concentration and global warming. In forecasting this time series, we may be able to predict what effects will we face from global warming.

```{r}
# Import Data
ts <- read.table("c02.txt", skip = 3, header = F, sep = "")

# Remove exraneous featurs
ts <- ts$V4

# Encode missing values
ts <- ifelse(ts == -99.99, NA, ts)

# Convert vector to time series object
ts <- ts(ts, start=c(1958,3), frequency=12)

# Plot the time series
plot.ts(ts)
mean(ts, na.rm=T)
abline(h=mean(ts, na.rm=T))
fit <- lm(ts$~seq(1,length(ts)))
plot(ts)
abline(fit)

```

This plot clearly indicates an upward trend with seasonality. This implies that we will need to perform differencing at lag 1 to remove the trend and differencing at lag 12 to remove seasonality. Since variance appears to stable, we need not perform a data transformation.

```{r}
# Remove trend
ts <- diff(ts, 1)
plot.ts(ts)

# Remove Seasonality
ts <- diff(ts, 12)
plot.ts(ts)
```

After differencing, the model appears stationary. We now examine ACF and PACF to preliminarily identify our model.

```{r}
acf <- acf(ts, lag = 50, type = "correlation", plot = TRUE, na.action = na.pass)
pacf <- pacf(ts, lag = 50, plot = TRUE, na.action = na.pass)
```

```{r}
arima(ts)
```

From these plots, we should consider $AR(1)$, $AR(2)$, $MA(1)$, and $ARMA(p,q)$ $p\leq2, \ q\leq1$ models.
